Commutative Character Sheaves

### 2015-10-27 followup.tex
Work from ICERM houseboat

Let $G$ be a smooth group scheme over a finite field $\Fq$; that is, let $G$ be a group scheme over $\Fq$
for which the structure morphism $G \to \Spec{\Fq}$ is smooth in the sense of [include reference for this definition].
This implies $G \to \Spec{\Fq}$ is locally of finite type, but not that it is of finite type.
We remark that the identity component $G^0$ of $G$ is of finite type over $\Fq$, while the component group scheme
$\pi_0(G)$ of $G$ is an \'etale group scheme over $\Fq$, and both are smooth over $\Fq$.

@@

@@

Let $\theta : G\times G \to G\times G$ be given by $\theta(g,h) = (h,g)$.
Write $m : G \times G \to G$ for the multiplication morphism
and define $m\op : G \times G \to G$ by $m\op = m \circ \theta$.
For any Weil local system $\cs{L}$ on $G$ there is an isomorphism
$\vartheta : \cs{L}\boxtimes\cs{L} \to \theta^*(\cs{L}\boxtimes\cs{L})$ given on stalks by the canonical map
$\gcs{L}_{g} \otimes \gcs{L}_{h} \to \gcs{L}_{h} \otimes \gcs{L}_{g}$.

When $G$ is commutative, recall \cite{cunningham-roe:13a}*{Def. 1.1} that a character sheaf on $G$ is a triple $(\gcs{L}, \mu, \phi)$,
where $\gcs{L}$ is a rank-one $\ell$-adic local system on $\bG$, 
$\mu : \bm^* \gcs{L} \to \LxL$ is an isomorphism
of sheaves on $\bG \times \bG$, 
and $\phi : \Frob{G}^* \gcs{L} \to \gcs{L}$ is an isomorphism of sheaves on $\bG$;
the triple $(\gcs{L}, \mu, \phi)$ must also satisfy compatibility conditions specified in \cite{cunningham-roe:13a}*{Def. 1.1}.
Writing $\cs{L}$ for the Weil local system $(\gcs{L}, \mu, \phi)$, these compatibility conditions are simply this: $\mu : m^*\cs{L} \to \cs{L}\boxtimes\cs{L}$ is an isomorphism of Weil local systems.
The category of character sheaves on $G$ is denoted by $\CS(G)$.

\begin{definition}\label{def:CCS}
A linear character sheaf on $G$ is a \emph{commutative character sheaf} if there is an isomorphism $\xi: m^* \cs{L} \to m\op^* \cs{L}$ of [linear character sheaves?] on $G\times G$ such that the following diagram commutes.
  \[
  \begin{tikzcd}[row sep=30]
   m^*\cs{L} \arrow{d}{\xi} \arrow{r}{\mu} & \cs{L}\boxtimes\cs{L} \arrow{d}{\vartheta}\\
   m\op^*\cs{L} \arrow{r}{{\theta}^*\mu} &  {\bar \theta}^*(\cs{L}\boxtimes\cs{L})
  \end{tikzcd}
  \]
We write $\CCS(G)$ for the full subcategory of $\CS(G)$ consisting of commutative character sheaves.
 \end{definition}

We will see that this condition is not trivial: even on commutative group schemes over $\Fq$, not every linear character sheaf is commutative; even on connected reductive algebraic groups over $\Fq$, not every linear character sheaf is commutative.

\begin{lemma}\label{lem:symmetry}
If the isomorphism $\xi: m^* \cs{L} \to m\op^* \cs{L}$ exists and satisfies the condition of Definition~\ref{def:CCS}, then it is unique.
\end{lemma}

\begin{proof}
Why is this true?
\end{proof}

@@

We now drop the assumption that $G$ is commutative, and assume only that it is smooth.
The category $\CS(G)$ still makes sense in this context, defined as in 
In order to distinguish the resulting objects from
the character sheaves of Lusztig, we will refer to the former as \emph{linear character sheaves} (to evoke the one-dimensional character sheaves of \cite{Kamgarpour})
and again write $\CS(G)$ for the resulting category, with morphisms defined as in \cite{cunningham-roe:13a}*{Def. 1.1}.

@@

\section{Descent to the Abelianization}

\subsection{Derived subgroup and abelianization}

Let $G$ be a smooth group scheme.
Let $G\der \hookrightarrow G$ denote the closed subgroup generated by the smooth morphism $G\times G \to G$ given by $(x,y)\mapsto xyx^{-1}y^{-1}$. Then $G \times G \to G\der$ is a dominant morphism \cite{SGA3}*{{VI}$_B$, 7.2}.
Moreover, $G$ is commutative if and only if $G\der$ is trivial.
%
While the subgroup $G\der$ is smooth, it need not be connected. 
The restriction of $G\times G\to G\der$ to identity components gives a dominant morphism $G^0\times G^0 \to G^0\der$ by [SGA3, Tome 6B, Expos\'e 7, Prop 7.8]

The quotient $G\ab\ceq G/G\der$ exists in schemes over $\Fq$ by SGA3, Expos\'e VI, Tome A, Th\'eor\`eme 3.2, which also shows that the quotient $G \to G\ab$ is a faithfully flat and a $G\der$-torsor in the fppf topology.
\[
\begin{tikzcd}
{} & G\times G \arrow{d}{} &&&\\
1 \arrow{r} & G\der \arrow{r}{} & G \arrow{r}{q} & G\ab \arrow{r} & 1
\end{tikzcd}
\]

>>>>

Let $G$ be a smooth group scheme over a finite field $\Fq$; that is, let $G$ be a group scheme over $\Fq$
for which the structure morphism $G \to \Spec{\Fq}$ is smooth in the sense of \todo{include reference for this definition}.
This implies $G \to \Spec{\Fq}$ is locally of finite type, but not that it is of finite type.
We remark that the identity component $G^0$ of $G$ is of finite type over $\Fq$, while the component group scheme
$\pi_0(G)$ of $G$ is an \'etale group scheme over $\Fq$, and both are smooth over $\Fq$.

@@

We write $m : G \times G \to G$ for the multiplication morphism, and $G(k)^*$ for $\Hom(G(k), \EEx)$.
Define $\theta : G\times G \to G\times G$ by $\theta(g,h) = (h,g)$.

@@

Suppose now that $G$ is commutative.
Recall \cite{cunningham-roe:13a}*{Def. 1.1} that a character sheaf on $G$ is a triple $(\gcs{L}, \mu, \phi)$,
where $\gcs{L}$ is a rank-one $\ell$-adic local system on $\bG$, 
$\mu : \bm^* \gcs{L} \to \LxL$ is an isomorphism
of sheaves on $\bG \times \bG$, 
and $\phi : \Frob{G}^* \gcs{L} \to \gcs{L}$ is an isomorphism of sheaves on $\bG$;
the triple $(\gcs{L}, \mu, \phi)$ must also satisfy compatibility conditions specified in \emph{loc. cite}.
The category of character sheaves on $G$ is denoted by $\CS(G)$.

Let $\cs{L}$ be a character sheaf on $G$.  Since $m = m \circ \theta$ in the commutative case,
there is a canonical isomorphism $\xi : m^* \cs{L} \to \theta^* m^* \cs{L}$.
There is also an isomorphism $\vartheta : \cs{L}\boxtimes\cs{L} \to \theta^*(\cs{L}\boxtimes\cs{L})$
given on stalks by the canonical map $\gcs{L}_{g} \otimes \gcs{L}_{h} \to \gcs{L}_{h} \otimes \gcs{L}_{g}$.

\begin{definition}\label{def:CCScom}
A character sheaf $(\cs{L}, \mu)$ on a smooth commutative group scheme $G$ is \emph{commutative}
if the following diagram of Weil sheaves on $G \times G$ commutes.
  \[
  \begin{tikzcd}[row sep=30]
   m^*\cs{L} \arrow{d}{\xi} \arrow{r}{\mu} & \cs{L}\boxtimes\cs{L} \arrow{d}{\vartheta}\\
   \theta^*(m^*\cs{L}) \arrow{r}{{\theta}^*\mu} &  \theta^*(\cs{L}\boxtimes\cs{L})
  \end{tikzcd}
  \]
We write $\CCS(G)$ for the full subcategory of $\CS(G)$ consisting of commutative character sheaves.
 \end{definition}

@@

We now drop the assumption that $G$ is commutative, and assume only that it is smooth.
The category $\CS(G)$ still makes sense in this context, defined as in \cite{cunningham-roe:13a}*{Def. 1.1}.
In order to distinguish the resulting objects from
the character sheaves of Lusztig, we will refer to the former as \emph{linear character sheaves}
(to evoke the one-dimensional character sheaves of \cite{Kamgarpour})
and again write $\CS(G)$ for the resulting category, with morphisms defined as in \emph{loc. cite}.

We recall some facts about the abelianization of a smooth group scheme $G$.
Let $G\der \hookrightarrow G$ denote the closed subgroup generated by the smooth morphism
$G\times G \to G$ given by $c: (x,y)\mapsto xyx^{-1}y^{-1}$. Then
$G \times G \to G\der$ is a dominant morphism \cite{SGA3}*{Expos\'e {VI}$_B$, \S 7.2}.
Moreover, $G$ is commutative if and only if $G\der$ is trivial.
When $G$ is not connected, $G\der$ need not be connected, though it is smooth.
The restriction of $G\times G\to G\der$ to identity components gives a dominant
morphism $G^0\times G^0 \to G^0\der$ \cite{SGA3}*{Expos\'e {VI}$_B$, Prop. 7.8}.

The quotient $G\ab\ceq G/G\der$ exists in schemes over $\Fq$, and the quotient $q : G \to G\ab$
is faithfully flat and a $G\der$-torsor in the fppf topology \cite{SGA3}*{Expos\'e {VI}$_A$, Thm. 3.2}.
We summarize the setup in a diagram.
\[
\begin{tikzcd}
{} & G\times G \arrow{d}{c} &&&\\
1 \arrow{r} & G\der \arrow{r}{} & G \arrow{r}{q} & G\ab \arrow{r} & 1
\end{tikzcd}
\]

If $\chi \in G(k)^*$ is a character, it must vanish on the derived subgroup $G\der(k)$.
Kamgarpour gives an example \cite{kamgarpour:09a}*{(1.1)} of a character sheaf
that does not vanish on $G\der$, defined by the extension
\[
1 \to \mu_n \to \SL_n \to \PGL_n \to 1.
\]
In order to obtain a relationship between character sheaves on $G$ and characters of $G(k)$,
he opts to give a different definition of commutator.  Since we already need to adapt our
notion of character sheaf, even in the commutative case, we instead add restrictions
to the definition of commutative character sheaf and leave the definition of $G\der$ unchanged.

The first step in this process is to rule out character sheaves as above.
The following proposition will be proven in Section \ref{sec:descent}.
\begin{proposition}
Suppose $G$ is a smooth group scheme and $\cs{L} \in \CS(G)$ is a linear character sheaf on $G$.
Then the restriction of $\cs{L}$ to $G\der$ is trivial if and only if $\cs{L} \cong q^*(\cs{L}\ab)$ for some
character sheaf $\cs{L}\ab$ on $G\ab$.
\end{proposition}

We may then define commutative character sheaves on $G$.

\begin{definition}\label{def:CCS}
A linear character sheaf $\cs{L}$ on a smooth group scheme $G$ is \emph{commutative} if
\[
\cs{L} \cong q^* \cs{L}\ab
\]
for a commutative character sheaf $\cs{L}\ab$ on $G\ab$.  As in the commutative case,
we write $\CCS(G)$ for the full subcategory of $\CS(G)$ consisting of commutative character sheaves.
\end{definition}

Note that we may also pull back the map $\xi\ab : m\ab^* \cs{L}\ab \to \theta\ab^*(m\ab^* \cs{L})$ from Definition \ref{def:CCScom}.
The following proposition will be proven in Section \ref{sec:descent}.

\begin{proposition}
A linear character sheaf $(\cs{L}, \mu)$ on a smooth group scheme $G$ is commutative if and only if
\begin{enumerate}
\item $\cs{L} \cong q^* \cs{L}\ab$ for some character sheaf $\cs{L}\ab$ on $G\ab$,
\item the following diagram of Weil sheaves on $G \times G$ commutes.
  \[
  \begin{tikzcd}[row sep=30]
   m^*\cs{L} \arrow{d}[swap]{(q\times q)^*\xi\ab} \arrow{r}{\mu} & \cs{L}\boxtimes\cs{L} \arrow{d}{\vartheta}\\
   \theta^*(m^*\cs{L}) \arrow{r}{{\theta}^*\mu} &  \theta^*(\cs{L}\boxtimes\cs{L})
  \end{tikzcd}
  \]
\end{enumerate}
\end{proposition}

@@

\section{Descent to the Abelianization} \label{sec:descent}
